\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-08 response.tex

\documentclass[12pt]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[margin=20mm]{geometry}
\usepackage{amsmath,color,amsthm,enumerate,amsfonts,mathtools}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{paralist}
\usepackage{stmaryrd}

\newcommand{\done}{\textcolor{blue}{Done.}}
\newcommand{\tba}{\textcolor{red}{To be addresssed.}}

@@

Page 3.  The sentence 'For each $vw \in E(G')$ there is a path $P$ in $G$
between $v$ and we of length at most $k+1$'' is correct but should be made
more precise.  For example, there could be a short path between $v$ and
$w$ which does not involve any new dummy vertices.  You mean to say that
the 'subdivided path' between $v$ and $w$ has length at most $k+1$.

\done

Page 3.  You should add the assumption that no three edges cross at
the same point (which you can assume by perturbing the embedding).
Otherwise, you will not get a $(k+1, 2)$-shortcut system.

$P$ is used twice in the statement of Theorem 3.  It is probably
safer (and necessary?) to use a different variable for the second
occurrence.

\textcolor{blue}{We now use $\mathcal{S}$ for shortcut systems.}

@@

Page 15. Regarding the proof of Theorem 11, if it really is the same
proof as the proof of Theorem 2, why not just prove Theorem 11 (from
which Theorem 2 follows as a special case)?

\textcolor{blue}{We have written the entire paper starting with simpler cases before moving onto the more general. This approach may use a few more words, but we think the paper becomes more accessible for more readers this way.}

Page 16.  At the end of Page 16, I do not see why 'every pair of edges
cross at most once.'  Without further assumptions, it seems as if two
paths in the path system could intersect several times.  Therefore, by
drawing 'each edge vw of G alongside $P_vw$ in $G_0$', there may be two
edges which cross several times.  It seems this can be fixed by
choosing the path system carefully by 'uncrossing' paths which
intersect at more than two internal vertices.

\textcolor{blue}{The referee is right here: two edges may cross more than once. We do not want to `uncross' since this does not maintain the bound on the number of crossings per edge. We have revised the proof avoiding the claim that every pair of edges cross at most once. The lemma itself has not changed. }

### 2022-03-08 k-planar.tex
updates

\author{Vida Dujmovi\'c%
        \thanks{School of Computer Science and Electrical Engineering,
                University of Ottawa, Ottawa, Canada (\texttt{vida.dujmovic@uottawa.ca}).
                Research supported by NSERC and the Ontario Ministry of Research and Innovation.},\,\,
        Pat Morin%
        \thanks{School of Computer Science, Carleton University, Ottawa, Canada (\texttt{morin@scs.carleton.ca}). Research  supported by NSERC and the Ontario Ministry of Research and Innovation.},\,\, and
        David R. Wood\thanks{School of Mathematics, Monash University, Melbourne, Australia (\texttt{david.wood@monash.edu}). Research supported by the Australian Research Council.}
}

@@

The starting point for this work is the following `product structure theorem' for planar graphs\footnote{In this paper, all graphs are finite and undirected. Unless specifically mentioned otherwise, all graphs are also simple. For any graph $G$ and any set $S$ (typically $S\subseteq V(G)$), let $G[S]$  denote the graph with vertex set $V(G)\cap S$ and edge set $\{uv\in E(G) : u,v\in S\}$.  We use $G-S$ as a shorthand for $G[V(G)\setminus S]$. We use $G'\subseteq G$ to denote subgraph containment; that is, $V(G')\subseteq V(G)$ and $E(G')\subseteq E(G)$.  We say that a graph $G$ is \defin{contained in} a graph $X$ if $G$ is isomorphic to a subgraph of $X$. Undefined terms are in \citep{Diestel5}.} by \citet{DJMMUW20} (improved by \citet{UWY}).

\begin{thm}[\citep{DJMMUW20,UWY}]
	\label{PlanarProduct}
	Every planar graph is contained in: 
	%a subgraph of:
	\begin{compactenum}[(a)]
		\item $H\boxtimes P$ for some graph $H$ of treewidth at most $8$ and for some path $P$,
		\item $H\boxtimes P \boxtimes K_3$ for some graph $H$ of treewidth at most $3$ and for some path $P$.
	\end{compactenum}
\end{thm}

% \note{DW}{In 	\cref{PlanarProduct} and elsewhere, we really should write, ``Every planar graph is isomorphic to a subgraph of''. One alternative is to define that $H$ is \defin{contained} in $G$ if $H$ is isomorphic to a subgraph of $G$. I suggest the latter.}

 \note{DW}{replace $8$ by $6$ and update elsewhere}

Here $\boxtimes$ is the strong product,\!\footnote{The \emph{strong product} of graphs $A$ and $B$, denoted by $A\boxtimes B$, is the graph with vertex set $V(A)\times V(B)$, where distinct vertices $(v,x),(w,y)\in V(A)\times V(B)$ are adjacent if
	$v=w$ and $xy\in E(B)$, or
	$x=y$ and $vw\in E(A)$, or
	$vw\in E(A)$ and $xy\in E(B)$.}
and treewidth\footnote{For a tree $T$, a \defin{$T$-decomposition} of a graph $G$ is a collection $\mathcal{T}=(B_x:x\in V(T))$ of subsets of $V(G)$ indexed by the nodes of $T$ such that
(i) for every $vw\in E(G)$, there exists some node $x\in V(T)$ with $v,w\in B_x$; and
(ii) for every $v\in V(G)$, the induced subgraph $T[v] := T[\{x: v\in B_x\}]$ is connected. The \defin{width} of $\mathcal{T}$ is $\max\{|B_x|:x\in V(T)\}-1$.  A \defin{tree-decomposition} is a $T$-decomposition for any tree $T$. The \defin{treewidth} $\tw(G)$ of a graph $G$ is the minimum width of a tree-decomposition of $G$.  Treewidth is the standard measure of how similar a graph is to a tree. Indeed, a connected graph has treewidth 1 if and only if it is a tree. Treewidth is of fundamental importance in structural and algorithmic graph theory; see \citep{Reed03,HW17,Bodlaender-TCS98} for surveys.} is an invariant that measures how `tree-like' a given graph is; see \cref{ProductExample} for an example. Loosely speaking, \cref{PlanarProduct} says that every planar graph is contained in the product of a tree-like graph and a path. This enables combinatorial results for graphs of bounded treewidth to be generalised for planar graphs (with different constants).

@@

\noindent\cref{PlanarProduct} has been the key tool in solving the following well-known open problems:
\begin{compactitem}
\item \citet{DJMMUW20} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92}).
\item  \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02}).
\item \citet{DFMS21} use it to make dramatic improvements to the best known bounds for $p$-centered colourings of planar graphs.
\item \citet{bonamy.gavoille.ea:shorter} use it to find shorter adjacency labellings of planar graphs (improving on a sequence of results starting with the work of \citet{kannan.naor.ea:implicit-stoc,kannan.naor.ea:implicit}). \citet{DEJGMM21} have since used it to find asymptotically optimal adjacency labellings of planar graphs.
\item The result of \citet{DEJGMM21} implies that, for every integer
$n>0$, there is a `induced-universal' graph $U_n$ with $n^{1+o(1)}$ vertices such that every $n$-vertex planar graph is an induced
subgraph of $U_n$. This has recently been extended by \citet{EJM}, who show the existence of a `subgraph-universal' graph with $(1+\epsilon)n$ vertices that contains every $n$-vertex planar graph as a subgraph as well as the existence of an `induced universal' graph with $n^{1+o(1)}$ vertices \emph{and} edges.  The former result is the first progress on subgraph-universal graphs for planar graphs since the $O(n^{3/2})$-vertex construction of \citet{babai.chung.ea:on}.
% aking the first progress on this problem since the work of \citet{babai.xx} in 1984.
\end{compactitem}

% \note{DW}{Delete `from 1992' and  `from 2007'? replace ``going back to 1988~\citep{kannan.naor.ea:implicit-stoc,kannan.naor.ea:implicit}'' by
	% ``starting with the work of \citet{kannan.naor.ea:implicit}''}
% \note{PM}{Was this a referee request, or just because we don't need to sell it anymore?  I'm fine with the change either way, but didn't implement it.}

% \note{DW}{Add \citep{EJM} to the story.}

@@

\begin{thm}
\label{kPlanarProduct}
Every $k$-planar graph is a subgraph of $H\boxtimes P\boxtimes K_{18k^2+48k+30}$, for some graph $H$ of treewidth $\binom{k+4}{3}-1$ and for some path $P$.
\end{thm}

@@

@@

\begin{obs}
\label{AddDummy}
Every $k$-planar graph is a subgraph of $G^\SS$ for some planar graph $G$ and some $(k+1,2)$-shortcut system $\SS$ for $G$.
\end{obs}

@@

\begin{thm}
\label{ShortcutProduct}
Let $G$ be a subgraph of $H\boxtimes P \boxtimes K_\ell$, for some graph $H$ of treewidth at most $t$ and for some path $P$.
Let $\SS$ be a $(k,d)$-shortcut system for $G$. Then $G^\SS$ is a subgraph of $J\boxtimes P\boxtimes K_{d\ell(k^3+3k)}$ for some graph $J$ of treewidth at most $\binom{k+t}{t}-1$ and some path $P$.
\end{thm}

@@

\begin{thm}
\label{kPowerBasic}
For every planar graph $G$ with maximum degree $\Delta$ and for every integer $k\geq 1$, $G^k$ is a subgraph of $H\boxtimes P\boxtimes K_{6k^2(k^2+3)\Delta^{k}}$ for some graph $H$ of treewidth at most $\binom{k+3}{3}-1$ and some path $P$.
\end{thm}

@@

\begin{thm}[\citep{DJMMUW20,UWY,DHHW}]
	\label{GenusProduct}
	Every graph of Euler genus $g$ is a subgraph of:
	\begin{compactenum}[(a)]
		\item $H  \boxtimes P$ for some graph $H$ of treewidth at most $2g+6$  and some path $P$.
		\item $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some graph $H$ of treewidth at most $3$ and for some path $P$.
	\end{compactenum}
\end{thm}

\note{DW}{I suggest we replace $2g+8$ by $2g+6$ here and replace 4 by 3, and cite \citep{DHHW,UWY}. This will improve the bounds and (more importantly) simplify the presentation of various results (such as \cref{StringPartition,PowerGenus}) which currently distinguish the $g=0$ case. }

\note{PM}{I agree with this suggestion.  I've only added the other two references here, and haven't changed anything elsewhere yet.}

@@

\begin{thm}[\citep{DJMMUW20}]
	\label{ApexMinorFree}
	For every apex graph $X$, there exists $c\in\mathbb{N}$ such that every $X$-minor-free graph is a subgraph of $H\boxtimes P$ for some graph $H$ with $\tw(H)\leq c$ and some path $P$.
\end{thm}

@@

\begin{thm}[\citep{DEMWW22}]
	\label{MinorFreeDegree}
	For every graph $X$ there exists $c\in\mathbb{N}$ such that for every $\Delta\in\mathbb{N}$, every $X$-minor-free graph $G$ with maximum degree at most $\Delta$ is a subgraph of $H\boxtimes P$ for some graph $H$ with $\tw(H) \leq c\Delta$ and for some path $P$.
\end{thm}

%The next three results are immediate corollaries of \cref{GenusProduct,ApexMinorFree,ShortcutProduct,MinorFreeDegree}.
%
%\begin{thm}
%Let $\SS$ be a $(k,d)$-shortcut system for a graph $G$ of Euler genus $g$. Then $G^\SS$ is a subgraph of $H\boxtimes P$ for some graph $H$ of treewidth at most $d(k^3+3k)\binom{k+2g+8}{2g+8}-1$ and for some path $P$.
%\end{thm}
%
%\begin{thm}
%For every apex graph $X$ and for all integers $k,d\geq 1$, there is an integer $c$ such that for every $X$-minor-free graph $G$ and for every $(k,d)$-shortcut system $\SS$ for $G$, $G^\SS\subseteq H\boxtimes P$ for some graph $H$ with $\tw(H)\leq c$ and for some path $P$.
%\end{thm}
%
%\begin{thm}
%For every graph $X$ and for all integers $k,\Delta\geq 1$, there is an integer $c$ such that for every $X$-minor-free graph $G$ with maximum degree $\Delta$ and for every $(k,\Delta)$-shortcut system $\SS$ for $G$, $G^\SS \subseteq H \boxtimes P$ for some graph $H$ with $\tw(H)\leq c$ and for some path $P$.
%\end{thm}
%
%\note{DW}{delete the above three theorems, they are never used}

@@

A \defin{layering} of a graph $G$ is a sequence $\mathcal{L}=\langle L_0,L_1,\ldots\rangle$ such that $\{L_0,L_1,\ldots\}$ is a partition of $V(G)$ and for every edge $vw\in E(G)$, if $v\in L_i$ and $w\in L_j$ then $|j-i|\leq 1$.  For any partition $\mathcal{P}=\{S_1,\ldots,S_p\}$ of $V(G)$, a \defin{quotient graph} $H=G/\mathcal{P}$ has a $p$-element vertex set $V(H)=\{x_1,\ldots,x_p\}$ and $x_ix_j\in E(H)$ if and only if there exists an edge $vw\in E(G)$ such that $v\in S_i$ and $w\in S_j$. To highlight the importance of the quotient graph $H$, we call $\mathcal{P}$ an \defin{$H$-partition} and write this concisely as $\mathcal{P}=\{S_x : x\in V(H)\}$ so that each element of $\mathcal{P}$ is indexed by the vertex it creates in $H$.

For any partition $\mathcal{P}$ of $V(G)$ and any layering $\mathcal{L}$ of $G$ we define the \defin{layered width} of $\mathcal{P}$ with respect to $\mathcal{L}$ as $\max\{|L\cap P|: L\in\mathcal{L},\, P\in\mathcal{P}\}$.  For any partition $\mathcal{P}$ of $V(G)$, we define the \defin{layered width} of $\mathcal{P}$ as the minimum, over all layerings $\mathcal{L}$ of $G$, of the layered width of $\mathcal{P}$ with respect to $\mathcal{L}$.

@@

\begin{lem}[\citep{DJMMUW20}]
\label{PartitionProduct}
For every graph $H$, a graph $G$ has an $H$-partition of layered width at most $\ell$ if and only if $G$ is a subgraph of $H \boxtimes P \boxtimes K_\ell$ for some path $P$.
\end{lem}

As an example of the use of layered partitions, to prove \cref{PlanarProduct}(a),
\citet{DJMMUW20} showed that every planar graph has an $H$-partition of layered width $1$ for some planar graph $H$ of treewidth at most $8$. The proof is constructive and gives a simple quadratic-time algorithm for finding the corresponding partition and layering.
At the core of their work is the elegant proof by \citet{PS21} of the following result:

\begin{thm}[\citep{PS21}]
\label{ps}
Every planar triangulation $G$ has an $H$-partition $\mathcal{P}$
such that $\tw(H)\leq 8$ and $G[P]$ is a shortest path in $G$ for each $P\in\mathcal{P}$.
\end{thm}

Indeed, the above-mentioned result of \citet{DJMMUW20} is a slight strengthening of \cref{ps}, where for each $P\in\mathcal{P}$ no two vertices of $P$ have the same distance to some fixed root vertex $r$.

@@

\note{DW}{Use $\SS$ for the given shortcut system and $\SS$ for the partition?}
\note{PM}{Yes, I've done this.  Again, I was carefuly, but need to reread.}

For each $x\in V(T)$, define $S_x := \{v\in V(G): a(v)= x\}$. Observe that $\mathcal{P}:=(S_x : x\in V(T))$ is a partition of $V(G)$.\footnote{The sets in $\mathcal{P}$ are disjoint because $a:V(G)\to V(T)$ is a function.  The sets in $\mathcal{P}$ cover $V(G)$ since, for each $v\in V(G)$, $v\in S_{a(v)}$.} Let $J:=G^\SS/\mathcal{P}$ denote the resulting quotient graph. We consider $V(J)\subseteq V(T)$, where each $x\in V(J)$ is the vertex obtained by contracting $S_x$ in $G^{\SS}$. (Any node $x\in V(T)$ with $S_x=\emptyset$ does not contribute a vertex to $J$.)

% \referee{2}{Page 8.  Can you add a brief explanation why $\mathcal{P}$ is a
% 	partition of $V(G)$?  It is clear that the sets in $\mathcal{P}$ are
% 	disjoint, but why do they cover $V(G)$?}
% \note{PM}{I added an explanation in a footnote since this is obvious.  In particular, for any function $f$, $\{f^{-1}(x):x\in\operatorname{range}(f)\}$ is a partition of $\operatorname{domain}(f)$.}

From this point onward, the plan is to show that:
\begin{compactenum}[(i)]
  \item $\mathcal{P}$ has small layered width with respect to the layering $\mathcal{L}$ of $G$ and that
  \item $J$ has small treewidth.
\end{compactenum}
Once we have established (i) and (ii), the result follows easily since a layering of $G^\SS$ is easily obtained from $\mathcal{L}$ by `compressing' groups of $k$ consecutive layers.
% \referee{1}{7. page 8, item (i): ``$P$ has small layered width with respect to the layering $L$'' $\longrightarrow$ ``$P$ has small layered width with respect to the layering $L$ of $G$''.}
% \referee{1}{Perhaps mention something about how it is easy to find the new layering.}

First we need to understand the relationship between sets $S_x\in\mathcal{P}$ and the related sets $Y_x$, $V_x$, $F_x$, and $N_x$.

@@

Next we complete Step~(i) and show that $\mathcal{P}$ has small layered width with respect to the layering $\mathcal{L}=\langle L_1,\ldots,L_h\rangle$:

@@

@@

We first formally define $k$-planar graphs.  An \defin{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a curve\footnote{A \defin{curve} in a surface $\Sigma$ is a continuous function $f:[0,1]\to \Sigma$. The points $f(0)$ and $f(1)$ are called the \defin{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior.  A \defin{crossing} in an embedded graph $G$ is a triple $(p,vw,xy)$ with $p\in\R^2$, $vw,xy\in E(G)$ and such that $p\in (vw\cap xy)\setminus\{v,w,x,y\}$. An embedded graph $G$ is \defin{$k$-plane} if each edge of $G$ takes part in at most $k$ crossings. \note{DW}{We need to assume that at most two edges cross at a single point, so we can add a dummy vertex of degree 4, right?}\note{PM}{No, under the current definition, $r$ edges crossing at a single point $p$ generate $\binom{r}{2}$ crossings, with each edge taking part in $r-1$ crossings, just like they would if we perturbed the edges so that each pair crosses at a distinct point.  I guess we could mention this.  (See the first paragraph in the proof of Theorem~2 below.)}  A (not necessarily embedded) graph $G'$ is \defin{$k$-planar} if there exists a $k$-plane graph $G$ isomorphic to $G'$. Under these definitions, $0$-planar graphs are exactly planar graphs and $0$-plane graphs are exactly plane graphs.

@@

\note{DW}{This should be fixed if we define kite more precisely. A kite-face is bounded by one edge and two half-edges right?}

>>>>

\author{%
Vida Dujmovi\'c%
\thanks{School of Computer Science and Electrical Engineering, University of Ottawa, Ottawa, Canada (\texttt{vida.dujmovic@uottawa.ca}). Research supported by NSERC and the Ontario Ministry of Research and Innovation.},\,\,
Pat Morin%
\thanks{School of Computer Science, Carleton University, Ottawa, Canada (\texttt{morin@scs.carleton.ca}). Research  supported by NSERC and the Ontario Ministry of Research and Innovation.},\,\, and
David R. Wood%
\thanks{School of Mathematics, Monash University, Melbourne, Australia (\texttt{david.wood@monash.edu}). Research supported by the Australian Research Council.}
}

@@

\note{DW}{I suggest we use consecutive numbering of theorems, lemmas, etc. Theorem 1, Lemma 2, Observation 3, etc.}

\note{DW}{I suggest we delete ``We use $G'\subseteq G$ to denote subgraph containment; that is, $V(G')\subseteq V(G)$ and $E(G')\subseteq E(G)$. '' This is absolutely standard.}

The starting point for this work is the following `product structure theorem' for planar graphs\footnote{In this paper, all graphs are finite and undirected. Unless mentioned otherwise, all graphs are also simple. For any graph $G$ and any set $S$ (typically $S\subseteq V(G)$), let $G[S]$  denote the graph with vertex set $V(G)\cap S$ and edge set $\{uv\in E(G) : u,v\in S\}$.  We use $G-S$ as a shorthand for $G[V(G)\setminus S]$. Undefined terms are in \citep{Diestel5}.} by \citet{DJMMUW20} (improved by \citet{UWY}). A graph $G$ is \defin{contained in} a graph $X$ if $G$ is isomorphic to a subgraph of $X$.

\begin{thm}[\citep{DJMMUW20,UWY}]
\label{PlanarProduct}
Every planar graph is contained in: 
\begin{compactenum}[(a)]
	\item $H\boxtimes P$ for some graph $H$ of treewidth at most $6$ and for some path $P$,
	\item $H\boxtimes P \boxtimes K_3$ for some graph $H$ of treewidth at most $3$ and for some path $P$.
\end{compactenum}
\end{thm}

Here $\boxtimes$ is the strong product,\!\footnote{The \defin{strong product} of graphs $A$ and $B$, denoted by $A\boxtimes B$, is the graph with vertex set $V(A)\times V(B)$, where distinct vertices $(v,x),(w,y)\in V(A)\times V(B)$ are adjacent if
	$v=w$ and $xy\in E(B)$, or
	$x=y$ and $vw\in E(A)$, or
	$vw\in E(A)$ and $xy\in E(B)$.}
and treewidth\footnote{For a tree $T$, a \defin{$T$-decomposition} of a graph $G$ is a collection $\mathcal{T}=(B_x:x\in V(T))$ of subsets of $V(G)$ indexed by the nodes of $T$ such that
(i) for every $vw\in E(G)$, there exists some node $x\in V(T)$ with $v,w\in B_x$; and
(ii) for every $v\in V(G)$, the induced subgraph $T[v] := T[\{x: v\in B_x\}]$ is connected. The \defin{width} of $\mathcal{T}$ is $\max\{|B_x|:x\in V(T)\}-1$.  A \defin{tree-decomposition} is a $T$-decomposition for any tree $T$. The \defin{treewidth} $\tw(G)$ of a graph $G$ is the minimum width of a tree-decomposition of $G$.  Treewidth is the standard measure of how similar a graph is to a tree. Indeed, a connected graph has treewidth 1 if and only if it is a tree. Treewidth is of fundamental importance in structural and algorithmic graph theory; see \citep{Reed03,HW17,Bodlaender-TCS98} for surveys.} is an invariant that measures how `tree-like' a given graph is; see \cref{ProductExample} for an example. Loosely speaking, \cref{PlanarProduct} says that every planar graph is contained in the product of a tree-like graph and a path. This enables combinatorial results for graphs of bounded treewidth to be generalised for planar graphs (with different constants).

@@

\noindent\cref{PlanarProduct} has been the key tool in solving the following well-known open problems:
\begin{compactitem}
\item \citet{DJMMUW20} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92}).
\item  \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02}).
\item \citet{DFMS21} use it to make dramatic improvements to the best known bounds for $p$-centered colourings of planar graphs.
\item \citet{bonamy.gavoille.ea:shorter} use it to find shorter adjacency labellings of planar graphs (improving on a sequence of results starting with the work of \citet{kannan.naor.ea:implicit}). \citet{DEJGMM21} have since used it to find asymptotically optimal adjacency labellings of planar graphs.
\item The result of \citet{DEJGMM21} implies that, for every integer
$n>0$, there is a `induced-universal' graph $U_n$ with $n^{1+o(1)}$ vertices such that every $n$-vertex planar graph is an induced
subgraph of $U_n$. This has recently been extended by \citet{EJM}, who show the existence of a `subgraph-universal' graph with $(1+\epsilon)n$ vertices that contains every $n$-vertex planar graph as a subgraph as well as the existence of an `induced universal' graph with $n^{1+o(1)}$ vertices \emph{and} edges.  The former result is the first progress on subgraph-universal graphs for planar graphs since the $O(n^{3/2})$-vertex construction of \citet{babai.chung.ea:on}.
% aking the first progress on this problem since the work of \citet{babai.xx} in 1984.
\end{compactitem}

\note{DW}{Mention other application such as vertex ranking~\citep{BDJM}, twin-width~\citep{BKW}, odd colourings~\citep{DMO}? We write a lot about adjacency labellings and universal induced subgraphs. Do we need so much detail here? We also mention this work in \cref{WrappingUp}.}

@@

\begin{thm}
\label{kPlanarProduct}
Every $k$-planar graph is contained in $H\boxtimes P\boxtimes K_{18k^2+48k+30}$, for some graph $H$ of treewidth $\binom{k+4}{3}-1$ and for some path $P$.
\end{thm}

@@

\note{DW}{I suggest we write, ``This theorem has applications in diverse areas, which we explore in \cref{Applications}....''}

@@

\begin{obs}
\label{AddDummy}
Every $k$-planar graph is contained in $G^\SS$ for some planar graph $G$ and some $(k+1,2)$-shortcut system $\SS$ for $G$.
\end{obs}

@@

\begin{thm}
\label{ShortcutProduct}
Let $G$ be a graph contained in $H\boxtimes P \boxtimes K_\ell$, for some graph $H$ of treewidth at most $t$ and for some path $P$. Let $\SS$ be a $(k,d)$-shortcut system for $G$. Then $G^\SS$ is contained in $J\boxtimes P\boxtimes K_{d\ell(k^3+3k)}$ for some graph $J$ of treewidth at most $\binom{k+t}{t}-1$ and some path $P$.
\end{thm}

@@

\begin{thm}
\label{kPowerBasic}
For every planar graph $G$ with maximum degree $\Delta$ and for every integer $k\geq 1$, $G^k$ is contained in $H\boxtimes P\boxtimes K_{6k^2(k^2+3)\Delta^{k}}$ for some graph $H$ of treewidth at most $\binom{k+3}{3}-1$ and some path $P$.
\end{thm}

@@

\begin{thm}[\citep{DJMMUW20,UWY,DHHW}]
\label{GenusProduct}
Every graph of Euler genus $g$ is contained in:
\begin{compactenum}[(a)]
\item $H  \boxtimes P$ for some graph $H$ of treewidth at most $2g+6$  and some path $P$.
\item $H \boxtimes P \boxtimes K_{\max\{2g,3\}}$ for some graph $H$ of treewidth at most $3$ and for some path $P$.
\end{compactenum}
\end{thm}

@@

\begin{thm}[\citep{DJMMUW20}]
	\label{ApexMinorFree}
	For every apex graph $X$, there exists $c\in\mathbb{N}$ such that every $X$-minor-free graph is contained in $H\boxtimes P$ for some graph $H$ with $\tw(H)\leq c$ and some path $P$.
\end{thm}

@@

\begin{thm}[\citep{DEMWW22}]
	\label{MinorFreeDegree}
	For every graph $X$ there exists $c\in\mathbb{N}$ such that for every $\Delta\in\mathbb{N}$, every $X$-minor-free graph $G$ with maximum degree at most $\Delta$ is contained in $H\boxtimes P$ for some graph $H$ with $\tw(H) \leq c\Delta$ and for some path $P$.
\end{thm}

%The next three results are immediate corollaries of \cref{GenusProduct,ApexMinorFree,ShortcutProduct,MinorFreeDegree}.
%
%\begin{thm}
%Let $\SS$ be a $(k,d)$-shortcut system for a graph $G$ of Euler genus $g$. Then $G^\SS$ is contained in $H\boxtimes P$ for some graph $H$ of treewidth at most $d(k^3+3k)\binom{k+2g+8}{2g+8}-1$ and for some path $P$.
%\end{thm}
%
%\begin{thm}
%For every apex graph $X$ and for all integers $k,d\geq 1$, there is an integer $c$ such that for every $X$-minor-free graph $G$ and for every $(k,d)$-shortcut system $\SS$ for $G$, $G^\SS\subseteq H\boxtimes P$ for some graph $H$ with $\tw(H)\leq c$ and for some path $P$.
%\end{thm}
%
%\begin{thm}
%For every graph $X$ and for all integers $k,\Delta\geq 1$, there is an integer $c$ such that for every $X$-minor-free graph $G$ with maximum degree $\Delta$ and for every $(k,\Delta)$-shortcut system $\SS$ for $G$, $G^\SS \subseteq H \boxtimes P$ for some graph $H$ with $\tw(H)\leq c$ and for some path $P$.
%\end{thm}
%
%\note{DW}{delete the above three theorems, they are never used}

@@

A \defin{layering} of a graph $G$ is a sequence $\mathcal{L}=\langle L_0,L_1,\ldots\rangle$ such that $\{L_0,L_1,\ldots\}$ is a partition of $V(G)$ and for every edge $vw\in E(G)$, if $v\in L_i$ and $w\in L_j$ then $|j-i|\leq 1$.  For any partition $\PP=\{S_1,\ldots,S_p\}$ of $V(G)$, a \defin{quotient graph} $H=G/\PP$ has a $p$-element vertex set $V(H)=\{x_1,\ldots,x_p\}$ and $x_ix_j\in E(H)$ if and only if there exists an edge $vw\in E(G)$ such that $v\in S_i$ and $w\in S_j$. To highlight the importance of the quotient graph $H$, we call $\PP$ an \defin{$H$-partition} and write this concisely as $\PP=\{S_x : x\in V(H)\}$ so that each element of $\PP$ is indexed by the vertex it creates in $H$.

For any partition $\PP$ of $V(G)$ and any layering $\mathcal{L}$ of $G$ we define the \defin{layered width} of $\PP$ with respect to $\mathcal{L}$ as $\max\{|L\cap P|: L\in\mathcal{L},\, P\in\PP\}$.  For any partition $\PP$ of $V(G)$, we define the \defin{layered width} of $\PP$ as the minimum, over all layerings $\mathcal{L}$ of $G$, of the layered width of $\PP$ with respect to $\mathcal{L}$.

@@

\begin{lem}[\citep{DJMMUW20}]
\label{PartitionProduct}
For every graph $H$, a graph $G$ has an $H$-partition of layered width at most $\ell$ if and only if $G$ is contained in $H \boxtimes P \boxtimes K_\ell$ for some path $P$.
\end{lem}

As an example of the use of layered partitions, to prove \cref{PlanarProduct}(a),
\citet{DJMMUW20} showed that every planar graph has an $H$-partition of layered width $1$ for some planar graph $H$ of treewidth at most $8$ (improved to $6$ in \cref{UWY}). The proof is constructive and gives a simple quadratic-time algorithm for finding the corresponding partition and layering. At the core of their work is the elegant proof by \citet{PS21} of the following result:

\begin{thm}[\citep{PS21}]
\label{ps}
Every planar triangulation $G$ has an $H$-partition $\PP$
such that $\tw(H)\leq 8$ and $G[P]$ is a shortest path in $G$ for each $P\in\PP$.
\end{thm}

Indeed, the above-mentioned result of \citet{DJMMUW20} is a slight strengthening of \cref{ps}, where for each $P\in\PP$ no two vertices of $P$ have the same distance to some fixed root vertex $r$.

@@

\note{DW}{Use $\SS$ for the given shortcut system and $\PP$ for the partition?}
\note{PM}{Yes, I've done this.  Again, I was carefuly, but need to reread.}

For each $x\in V(T)$, define $S_x := \{v\in V(G): a(v)= x\}$. Observe that $\PP:=(S_x : x\in V(T))$ is a partition of $V(G)$.\footnote{The sets in $\PP$ are disjoint because $a:V(G)\to V(T)$ is a function.  The sets in $\PP$ cover $V(G)$ since, for each $v\in V(G)$, $v\in S_{a(v)}$.} Let $J:=G^\SS/\PP$ denote the resulting quotient graph. We consider $V(J)\subseteq V(T)$, where each $x\in V(J)$ is the vertex obtained by contracting $S_x$ in $G^{\SS}$. (Any node $x\in V(T)$ with $S_x=\emptyset$ does not contribute a vertex to $J$.)

% \referee{2}{Page 8.  Can you add a brief explanation why $\PP$ is a
% 	partition of $V(G)$?  It is clear that the sets in $\PP$ are
% 	disjoint, but why do they cover $V(G)$?}
% \note{PM}{I added an explanation in a footnote since this is obvious.  In particular, for any function $f$, $\{f^{-1}(x):x\in\operatorname{range}(f)\}$ is a partition of $\operatorname{domain}(f)$.}

From this point onward, the plan is to show that:
\begin{compactenum}[(i)]
  \item $\PP$ has small layered width with respect to the layering $\mathcal{L}$ of $G$ and that
  \item $J$ has small treewidth.
\end{compactenum}
Once we have established (i) and (ii), the result follows easily since a layering of $G^\SS$ is easily obtained from $\mathcal{L}$ by `compressing' groups of $k$ consecutive layers.
% \referee{1}{7. page 8, item (i): ``$P$ has small layered width with respect to the layering $L$'' $\longrightarrow$ ``$P$ has small layered width with respect to the layering $L$ of $G$''.}
% \referee{1}{Perhaps mention something about how it is easy to find the new layering.}

First we need to understand the relationship between sets $S_x\in\PP$ and the related sets $Y_x$, $V_x$, $F_x$, and $N_x$.

@@

Next we complete Step~(i) and show that $\PP$ has small layered width with respect to the layering $\mathcal{L}=\langle L_1,\ldots,L_h\rangle$:

@@

  \note{DW}{replace Lemma~24 by Lemma~13, right?}

@@

We first formally define $k$-planar graphs.  An \defin{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a curve\footnote{A \defin{curve} in a surface $\Sigma$ is a continuous function $f:[0,1]\to \Sigma$. The points $f(0)$ and $f(1)$ are called the \defin{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior.  A \defin{crossing} in an embedded graph $G$ is a triple $(p,vw,xy)$ with $p\in\R^2$, $vw,xy\in E(G)$ and such that $p\in (vw\cap xy)\setminus\{v,w,x,y\}$. An embedded graph $G$ is \defin{$k$-plane} if each edge of $G$ takes part in at most $k$ crossings. \note{DW}{We need to assume that at most two edges cross at a single point, so we can add a dummy vertex of degree 4, right?}\note{PM}{No, under the current definition, $r$ edges crossing at a single point $p$ generate $\binom{r}{2}$ crossings, with each edge taking part in $r-1$ crossings, just like they would if we perturbed the edges so that each pair crosses at a distinct point.  I guess we could mention this.  (See the first paragraph in the proof of Theorem~2 below.)} \note{DW}{The downside of this approach is that in the opening section we give an informal definition of $k$-planar, which needs the assumption of having no three edges cross at a point, since we use this assumption in the construction of a $(k+1,2)$-shortcut system for $k$-planar graphs give in the Introduction. I think we should not have different definitions in two parts of the paper.} A (not necessarily embedded) graph $G'$ is \defin{$k$-planar} if there exists a $k$-plane graph $G$ isomorphic to $G'$. Under these definitions, $0$-planar graphs are exactly planar graphs and $0$-plane graphs are exactly plane graphs.

@@

%\note{DW}{This should be fixed if we define kite more precisely. A kite-face is bounded by one edge and two half-edges right?}
